\chapter{The First-Order Theory of Hereditarily Finite Sets}

In this chapter, we introduce the first-order theory of hereditarily finite sets (HF).
Following \cite{shoenfield1967mathematical}, a first-order theory is a formal system
consisting of a first-order language, a set of non-logical axioms, and a logical calculus 
that includes logical axioms and deduction rules.
In fact, to specify the first-order theory of HF, it suffices to specify 
the non-logical symbols of the language and the non-logical axioms, as everything else is
given by the definition of a first-order theory.
However, as we aim to formalise the notion of a formal syntactical proof within the theory of HF,
we do present the logical axioms and deduction rules.
Furthermore, we need a description of the logical axioms and deduction rules for 
constructing the provability formula in Chapter 5.

For the formalisation of the first-order theory of HF,
we draw on formalised concepts from first-order logic available in 
the model theory section of the math library of Lean 4.
These formalised concepts, such as the notions of a first-order language,
a first-order structure and a first-order formula, are descendants from 
the Flypitch project by \cite{han2020formal}.

\section{Language and axioms}

The \textit{first-order theory of HF} is determined by the first-order language of HF and 
the (non-logical) axioms of HF.

\begin{definition}[The language of HF]
    \label{def:Lang}
    \lean{HF.Lang}
    \leanok
    The \textit{first-order language of HF}, denoted by $\mathcal{L}$, 
    has an alphabet of \textit{symbols} given by the 
    logical constants $=$, $\lor$, $\neg$, $\exists$, the variables denoted by 
    lower case Latin letters, one binary relation symbol $\in$, one binary function symbol $\lhd$, 
    and one constant symbol $\varnothing$.
    The connectives $\land$, $\rightarrow$, $\leftrightarrow$, and $\forall$ are considered 
    abbreviations.
\end{definition}

It follows that any \textit{term} of $\mathcal{L}$ arises in the following way:
any variable is a term, any constant symbol is a term, and if $\tau_1, \tau_2$ are terms, then so is
$\tau_1 \lhd \tau_2$.
Furthermore, any \textit{$\mathcal{L}$-formula} arises in the following way:
any atomic formula is an $\mathcal{L}$-formula, and if $\varphi, \psi$ are $\mathcal{L}$-formulas, 
then so are $\neg \varphi$, $\varphi \lor \psi$, and $\exists x \varphi$.
Again, $\varphi \land \psi$, $\varphi \rightarrow \psi$, $\varphi \leftrightarrow \psi$ and
$\forall x \varphi$ are treated as definitions.

The axioms of HF consist of two axioms and one scheme. 
Informally, the first axiom defines $\varnothing$ as the
empty set, the second axiom defines $\lhd$ as the operation of adjoining an element to a set,
and the axiom scheme describes proofs by induction.

\begin{definition}[The axioms of HF]
    \label{def:Axioms}
    \lean{HF.Axiom1, HF.Axiom2, HF.Axiom3}
    \uses{def:Lang}
    \leanok
    The \textit{axioms of HF} are the following $\mathcal{L}$-formulas:
    \begin{enumerate}
        \item $z=\varnothing \leftrightarrow \forall x(x \notin z)$.
        \item $z=x \lhd y \leftrightarrow \forall u(u \in z \leftrightarrow u \in x \lor u=y)$.
        \item $(\alpha(\varnothing) \land \forall x \forall y[\alpha(x) \land \alpha(y) \rightarrow 
        \alpha(x \lhd y)]) \rightarrow \forall x \alpha(x)$,
    \end{enumerate}
    where $\alpha$ is any $\mathcal{L}$-formula which contains a 
    distinguished freely occurring variable $z$ such that $x$ and $y$ are substitutable for $z$.
\end{definition}

\section{Logical calculus}

This section recalls the logical axioms and deduction rules for a first-order theory, 
as applied to HF. 
It can be readily shown that the \textit{logical calculus of HF}, as described below, 
is equivalent to the logical calculus of a first-order theory in \cite{shoenfield1967mathematical} 
(when applied to HF), in the sense that they produce the same theorems.

The \textit{logical axioms} are the following $\mathcal{L}$-formulas:

\begin{definition}[Sentential (Boolean) axioms]
    \label{def:Bool.Axioms}
    \lean{HF.Bool.Axiom1, HF.Bool.Axiom2, HF.Bool.Axiom3, HF.Bool.Axiom4, HF.Bool.Axiom5}
    \uses{def:Lang}
    \leanok
    For any $\mathcal{L}$-formulas $\varphi, \psi, \mu$:
    $$
    \begin{aligned}
    \varphi & \rightarrow \varphi, \\
    \varphi & \rightarrow \varphi \lor \psi, \\
    \varphi \lor \varphi & \rightarrow \varphi, \\
    \varphi \lor(\psi \lor \mu) & \rightarrow(\varphi \lor \psi) \lor \mu, \\
    (\varphi \lor \psi) \land(\neg \varphi \lor \mu) & \rightarrow \psi \lor \mu .
    \end{aligned}
    $$
\end{definition}

\begin{definition}[Specialisation axiom]
    \label{def:Spec.Axiom}
    \lean{HF.Spec.Axiom}
    \uses{def:Lang}
    \leanok
    For any $\mathcal{L}$-formula $\varphi$ and every $x_i$:
    $$\varphi \rightarrow \exists x_i \varphi.$$
\end{definition}

\begin{definition}[Equality axioms]
    \label{def:Equality.Axioms}
    \lean{HF.Equality.Axiom1, HF.Equality.Axiom2, HF.Equality.Axiom3, HF.Equality.Axiom4}
    \uses{def:Lang}
    \leanok
    $$
    \begin{aligned}
        x_1 & = x_1, \\
        \left(x_1=x_2\right) \land \left(x_3=x_4\right) & 
        \rightarrow \left[\left(x_1=x_3\right) \rightarrow \left(x_2=x_4\right)\right], \\
        \left(x_1=x_2\right) \land \left(x_3=x_4\right) & 
        \rightarrow \left[\left(x_1 \in x_3\right) \rightarrow \left(x_2 \in x_4\right)\right], \\
        \left(x_1=x_2\right) \land \left(x_3=x_4\right) & 
        \rightarrow \left[x_1 \lhd x_3 = x_2 \lhd x_4 \right]. 
    \end{aligned}
    $$
\end{definition}

The \textit{deduction rules} are as follows:

\begin{definition}[Deduction rules]
    \label{def:prf.MP+prf.Subst+prf.ExIntro}
    \lean{HF.prf.MP}
    \uses{def:Lang}
    For any $\mathcal{L}$-formulas $\varphi, \psi$:
    \begin{description}
        \item \textit{Modus Ponens}: 
        From $\varphi$ and $\varphi \rightarrow \psi$, deduce $\psi$.
        \item \textit{Substitution}:
        From $\varphi$, deduce $\varphi[x_i/\tau]$ for any term $\tau$ that is substitutable for
        $x_i$ in $\varphi$.
        \item \textit{$\exists$-introduction}:
        From $\varphi \rightarrow \psi$, deduce $\exists x_i \varphi \rightarrow \psi$ provided
        $x_i$ does not occur freely in $\psi$.
    \end{description}
\end{definition}

These descriptions naturally lead to the notion of formal proof.

\begin{definition}[Theorem of HF]
    \label{def:prf}
    \lean{HF.prf}
    \uses{def:Lang, def:Axioms, def:Bool.Axioms, def:Spec.Axiom, 
        def:Equality.Axioms, def:prf.MP+prf.Subst+prf.ExIntro}
    An $\mathcal{L}$-formula is a \textit{theorem of HF}, denoted by $\vdash \varphi$, 
    if there is a finite sequence of $\mathcal{L}$-formulas — each of which is an axiom of HF, 
    a logical axiom, or is deduced from previous formulas in the sequence by a deduction rule —
    terminating with $\varphi$.
\end{definition}

Following the approach outlined by \cite{paulson2014machine}, 
the syntactic proof system of HF is implemented in Lean as an inductively defined predicate. 
In this setup, a proof within this system, resulting in a theorem of HF, 
involves demonstrating in Lean’s native logic that 
the proof predicate applied to the relevant $\mathcal{L}$-formula evaluates to true.
However, in practice, working directly within this formal calculus is cumbersome. 
\cite{paulson2014machine} experienced that even the simplest formal proofs can become extremely 
tedious and require hundreds of lines of code.
To streamline the process of obtaining theorems of HF in Lean, 
we aim to convert semantic proofs into proofs within the HF syntactic system. 
This is possible by employing Gödel's completeness theorem, which applies to any first-order theory.
How this mechanism is put into practice is discussed in Chapter 2.

\section{Semantics}

Apart from drawing on the concept of a standard model, as introduced in Chapter 3, 
Gödel's incompleteness theorems are fundamentally syntactical in nature. 
They focus solely on what can be proven within the confines of the formal system itself, 
without relying on any external concepts like semantic notions.

However, while we do explain what it means to be a model to introduce the standard model of 
HF in Chapter 3, we also present the semantic results of soundness and completeness. 
This is done solely because these results are helpful in the process of converting semantic proofs 
into syntactic proofs in Lean, as discussed earlier.
Regarding all proofs presented in the coming chapters, we follow 
\cite{swierczkowski2003finite} — who does not introduce Gödel's completeness theorem — 
in terms of referring to semantic notions.

\begin{definition}
    \label{def:models+Model+valid}
    \lean{HF.models, HF.Model, HF.valid}
    \uses{def:Lang, def:Axioms}
    \leanok
    Suppose $\mathcal{S}$ is an $\mathcal{L}$-structure and $\varphi$ is an $\mathcal{L}$-formula.
    If every valuation in $\mathcal{S}$ satisfies $\varphi$, 
    say $\mathcal{S}$ is a \textit{model} of $\varphi$, denoted by $\mathcal{S} \vDash \varphi$.

    If $\mathcal{S}$ is a model of every axiom of HF, say $\mathcal{S}$ is a \textit{model of HF}.
    
    An $\mathcal{L}$-formula $\varphi$ is called \textit{valid in HF} if 
    every model of HF is a model of $\varphi$, denoted by $\vDash \varphi$.
\end{definition}

For a proof of the following result for an arbitrary first-order theory, 
the reader is referred to  \cite{shoenfield1967mathematical}.

\begin{theorem}[Soundness and completeness]
    \label{thm:completeness}
    \lean{HF.completeness}
    \uses{def:Lang, def:prf, def:models+Model+valid}
    \leanok
    Suppose $\varphi$ is an $\mathcal{L}$-formula.
    Then $\vdash \varphi$ if and only if $\vDash \varphi$.
\end{theorem}

In other words, the theorems of HF are precisely the $\mathcal{L}$-formulas valid in HF.
Thus, to prove $\vdash \varphi$ for any $\mathcal{L}$-formula $\varphi$, it suffices to show that for an 
arbitrary model of HF, $\mathcal{S}$, we have $\mathcal{S} \vDash \varphi$.
This is the conversion process we previously discussed.

In Lean 3, the completeness theorem for first-order logic was proven as part of the Flypitch project
by \cite{han2020formal}.
However, no proof is available in the math library of Lean 4.
For now, we make use of the extraneous concepts of soundness and completeness 
by accepting the statement using the {\ttfamily \small sorry} identifier. 